---
title: "Titanic Survival Prediction Model"
author: "Trevor Okinda"
date: ""
output:
  github_document: 
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
editor_options:
  chunk_output_type: console
---

# Student Details

|                                              |     |
|----------------------------------------------|-----|
| **Student ID Number**                        | 134780 |
| **Student Name**                             | Trevor Okinda |
| **BBIT 4.2 Group**                           | C |
| **Project Name**                             | Titanic Survival Prediction Model |

# Setup Chunk

**Note:** the following KnitR options have been set as the global defaults: <BR> `knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE, collapse = FALSE, tidy = TRUE)`.

More KnitR options are documented here <https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html> and here <https://yihui.org/knitr/options/>.

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = FALSE
)
```

# Understanding the Dataset (Exploratory Data Analysis (EDA))

## Loading the Dataset

### Source: 

The dataset that was used can be downloaded here: *\<https://www.kaggle.com/datasets/rahulsah06/titanic?resource=download\>*

### Reference:

*\<Avagyan, Z. (2017). Weather CSV [Data set]. Kaggle. https://www.kaggle.com/datasets/zaraavagyan/weathercsv\>\
Refer to the APA 7th edition manual for rules on how to cite datasets: <https://apastyle.apa.org/style-grammar-guidelines/references/examples/data-set-references>*

```{r Load dataset}
#Load dataset
TitanicData <-read.csv("titanic.csv", colClasses = c(
  PassengerId = "numeric",
  Pclass = "numeric",
  Name = "character",
  Sex = "factor",
  Age = "numeric",
  SibSp = "numeric",
  Parch = "numeric",
  Ticket = "character",
  Fare = "numeric",
  Cabin = "character",
  Embarked = "factor"
))

# Define levels for categorical columns
embarked_levels <- c("C", "Q", "S")
sex_levels <- c("female", "male")

# Update factor columns with levels
TitanicData$Embarked <- factor(TitanicData$Embarked, levels = embarked_levels)
TitanicData$Sex <- factor(TitanicData$Sex, levels = sex_levels)

# Display the dataset
View(TitanicData)
```

```{r Measures of Frequency}
# Display frequency and percentage of Pclass
table_Pclass <- table(TitanicData$Pclass)
percentage_Pclass <- prop.table(table_Pclass) * 100
class_labels_Pclass <- c("First Class", "Second Class", "Third Class")
cat("Frequency and Percentage of Pclass:\n")
cat(table_Pclass, "\n")
cat(paste0("Percentage:\n", round(percentage_Pclass, 2)), "\n")
cat(paste0("Class Labels:\n", class_labels_Pclass, "\n\n"))

# Display frequency and percentage of Sex
table_Sex <- table(TitanicData$Sex)
percentage_Sex <- prop.table(table_Sex) * 100
class_labels_Sex <- c("Female", "Male")
cat("Frequency and Percentage of Sex:\n")
cat(table_Sex, "\n")
cat(paste0("Percentage:\n", round(percentage_Sex, 2)), "\n")
cat(paste0("Class Labels:\n", class_labels_Sex, "\n\n"))

# Display frequency and percentage of Embarked
table_Embarked <- table(TitanicData$Embarked)
percentage_Embarked <- prop.table(table_Embarked) * 100
class_labels_Embarked <- c("Cherbourg", "Queenstown", "Southampton")
cat("Frequency and Percentage of Embarked:\n")
cat(table_Embarked, "\n")
cat(paste0("Percentage:\n", round(percentage_Embarked, 2)), "\n")
cat(paste0("Class Labels:\n", class_labels_Embarked, "\n\n"))

```

```{r Measures of Central Tendency}
# Calculate mean, median, and mode for Age
mean_Age <- mean(TitanicData$Age, na.rm = TRUE)
median_Age <- median(TitanicData$Age, na.rm = TRUE)
mode_Age <- as.numeric(names(table(TitanicData$Age))[which.max(table(TitanicData$Age))])

cat("Measures of Central Tendency for Age:\n")
cat(paste0("Mean: ", round(mean_Age, 2)), "\n")
cat(paste0("Median: ", median_Age), "\n")
cat(paste0("Mode: ", mode_Age), "\n\n")

# Calculate mean, median, and mode for Fare
mean_Fare <- mean(TitanicData$Fare, na.rm = TRUE)
median_Fare <- median(TitanicData$Fare, na.rm = TRUE)
mode_Fare <- as.numeric(names(table(TitanicData$Fare))[which.max(table(TitanicData$Fare))])

cat("Measures of Central Tendency for Fare:\n")
cat(paste0("Mean: ", round(mean_Fare, 2)), "\n")
cat(paste0("Median: ", median_Fare), "\n")
cat(paste0("Mode: ", mode_Fare), "\n\n")
```

```{r Measures of Distribution}
# Calculate range, variance, and standard deviation for Age
range_Age <- range(TitanicData$Age, na.rm = TRUE)
variance_Age <- var(TitanicData$Age, na.rm = TRUE)
sd_Age <- sd(TitanicData$Age, na.rm = TRUE)

cat("Measures of Distribution for Age:\n")
cat(paste0("Range: ", paste(range_Age, collapse = " - ")), "\n")
cat(paste0("Variance: ", round(variance_Age, 2)), "\n")
cat(paste0("Standard Deviation: ", round(sd_Age, 2)), "\n\n")

# Calculate range, variance, and standard deviation for Fare
range_Fare <- range(TitanicData$Fare, na.rm = TRUE)
variance_Fare <- var(TitanicData$Fare, na.rm = TRUE)
sd_Fare <- sd(TitanicData$Fare, na.rm = TRUE)

cat("Measures of Distribution for Fare:\n")
cat(paste0("Range: ", paste(range_Fare, collapse = " - ")), "\n")
cat(paste0("Variance: ", round(variance_Fare, 2)), "\n")
cat(paste0("Standard Deviation: ", round(sd_Fare, 2)), "\n\n")

```

```{r Measures of relationship}
# Calculate correlation between Age and Fare
correlation_Age_Fare <- cor(TitanicData$Age, TitanicData$Fare, use = "complete.obs")

cat("Measures of Relationship (Correlation) between Age and Fare:\n")
cat(paste0("Correlation Coefficient: ", round(correlation_Age_Fare, 2)), "\n\n")
```

